---
title: "Bern25 summary charts"
format: html
---
  <!-- html: -->
  <!--   page-layout: full -->

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(fs)
library(glue)
library(gt)

### Changelog:
# 260311  - initial version 
# 260315  - almost complete report 
# 260316  - moved to dedicated project 
```


## The scope of the reporting

```{r}
#| label: read_data
#| include: false

ninaServer <- F
pdrive <- if(ninaServer) "~/Mounts/P-Prosjekter2/112549_bern_2025" else "P:/112549_bern_2025"
ahome <-  "assessment" %>% path(pdrive, .) # working dir for the experts
tpldir0 <- "data/_templates" %>% path(pdrive, .) # working dir for templates
sthome <- "data/st_all" %>% path(pdrive, .) # a repo folder for the "full" contents of all "simple templates" (st)
jsonhome <- "../Bern25-harvest/out/" # the home of the json files harvested from the simple templates (...does not work on NINA server!!!)

### introduction of typologies, etc.
fgrps_en <- c(ma= "Mammals", 
              fa= "Fishes & amphibians",
              im= "Insects and molluscs",
              vp= "Vascular plants",
              mo= "Mosses",
              ht= "Habitat types",
              all="All")
para0_en <- c(Ran="Range (Ran)", 
              Pop="Population (Pop)", 
              Are="Area of habitats (Are)",
              H4s="Habitat for species (H4s)", 
              SnF="Structure and function (SnF)",
              Fpr="Future prospects (Fpr)",
              OvC="Overall conclusion (OvC)")


### helpers
h_lkp <- function(x, tab) { # generic lookup (x: any vec; tab: a df or mat w cols 1/2 containing lkp keys/values)
  tab[[2]][match(as.character(x), as.character(tab[[1]]))]
  } 
h_cap1st <- function(x) x %>% #capitalise just the first letters of a chr vector (all subsequent letters kept)
  str_sub(1,1) %>% toupper %>% str_c(str_sub(x,2)) 
h_ensure_cols <- function(x, cols, typ="") { # make sure that columns _cols_ exist in _x_ 
  new_cols <- cols[!cols %in% names(x)]      # (create them with NAs (of the type of typ) if needed)
  ret <- typ %>% c(NA, .) %>% {.[1]} %>% 
    rep(length(new_cols)) %>% set_names(new_cols) %>% 
    as.list %>% bind_cols(x, .)
  ret <- if (nrow(x)!=0L) ret else slice(ret, -1)
  ret}

### the checklist, xt reflovs and st lookups
chklw <- read_rds(path(pdrive,"data/chklist-newest.rds"))   #wide
chkll <- read_rds(path(pdrive,"data/chkl_long-newest.rds")) #long
reflovs <- read_csv(path(tpldir0, "xt_reflovs-260126.csv")) %>% rename(id= rd_id)
stcodes <- read_csv(path(tpldir0, "st_codelists-newest.csv"))
pmcodes <- read_csv(path(tpldir0, "pres_meas_codes-newest.csv"))

### lists of all extracted fields from the last versions of the two xlsx templates
tmp <- read_rds(path(tpldir0, "t_struc.rds")) # t_struc
fds <- NULL
fds$sp <- tmp %>% keep_at(\(i) str_detect(i,"^sp")) %>% pluck(-1) %>% select(-icell) 
fds$ha <- tmp %>% keep_at(\(i) str_detect(i,"^ha")) %>% pluck(-1) %>% select(-icell) 

### the i2019 "old values"
i2019all <- read_rds(path(pdrive,"data","2019_data","all_data_2019-01-250923_2008.rds"))

### the json files "harvested" from the simple templates
tmp <- chklw %>%
  with(str_c("b25_", exp0,"_", ftid, ".json")) %>% 
  path(jsonhome, .) %>%
  map(\(x) jsonlite::read_json(here::here(x)))

tmp1 <- tibble(ftid=  map_chr(tmp, \(x) pluck(x, 1, "ftid"))) %>%
  # mutate(ft_name0=   map_chr(tmp, \(x) pluck(x, 1, "ft_name0"))) %>%
  mutate(t_ver=   map_chr(tmp, \(x) pluck(x, 1, "t_ver"))) %>%
  mutate(tabs=    map(tmp, \(x) pluck(x, 1, "tabs"))) %>%
  mutate(tabs= map(tabs, \(x) list_transpose(x, default=NA))) %>%
  mutate(tabs= map(tabs, as_tibble)) 

### two full "data cubes" (for sp & ha separately) 
dat <- NULL # the two "data cubes" -- with everything (except for ranges, etc.)
dat$sp <- tmp1 %>% filter(str_detect(t_ver, "^sp")) %>% unnest(tabs) %>%
  bind_rows(set_names(fds$sp$f_id) %>% as.list %>% as_tibble %>% slice(-1), .) %>% #fix col order & explicit NAs
  left_join(chkll, by=join_by(ftid, bgr)) %>%
  left_join(select(chklw, -any_of(names(select(chkll, -ftid)))), by=join_by(ftid)) %>%
  mutate(fgrp= fct(fgrp, names(fgrps_en))) %>%
  relocate(ftid, bgr, fgrp, exp0, occ1, ft_name0, status)

dat$ha <- tmp1 %>% filter(str_detect(t_ver, "^ha")) %>% unnest(tabs) %>%
  bind_rows(set_names(fds$ha$f_id) %>% as.list %>% as_tibble %>% slice(-1), .) %>% #fix col order & explicit NAs
  left_join(chkll, by=join_by(ftid, bgr)) %>%
  left_join(select(chklw, -any_of(names(select(chkll, -ftid)))), by=join_by(ftid)) %>%
  mutate(fgrp= fct(fgrp, names(fgrps_en))) %>%
  relocate(ftid, bgr, fgrp, exp0, occ1, ft_name0, status)
```

There were altogether 72 *species* and 6 *habitat types* involved in the assessment (@tbl-fgrp). Species and habitat types are collectively called *features* so there were altogether 78 features assessed. (*TODO: explain BGRs, assessments (feature x BGR), Emerald, etc...*) 

```{r}
#| label: tbl-fgrp
#| tbl-cap: "Groups of features (species, habitats) used in this report (N~features~: number of features, N~assmt.full~: number of full assessments, N~assmt.part~: number of partial assessments)"
#| warning: false
#| echo: false

### an overview table
tmp1 <- chkll %>%
  mutate(fgrp= h_lkp(ftid, select(chklw, ftid, fgrp)) %>% fct(names(fgrps_en))) %>%
  group_by(fgrp) %>% summarise(Nf= length(unique(ftid)), Na1= sum(t_ass=="full"), Na2= n()-Na1) %>%
  mutate(main= str_c(fgrps_en[as.character(fgrp)]," (",fgrp,")") %>% unname) %>% 
  select(main, Nf, Na1, Na2)
tmp2 <- tmp1 %>% summarise(main= "Total", Nf=sum(Nf), Na1=sum(Na1), Na2=sum(Na2))
  
bind_rows(tmp1, tmp2) %>% gt() %>% 
  cols_label(main="Feature groups", Nf=md("N~features~"), Na1=md("N~assmt.full~"), Na2=md("N~assmt.part~")) %>%
  cols_align("center") %>% cols_align("left", columns= main) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```


## Conservation status of species and habitat types 

We start with presenting an overview of the conclusions of the assessments. Each assessment has an overall conclusion, which is based on four "parameters", each of which has its own conclusions. The conclusions for the individual parameters rely on quantitative and qualitative data collected by taxonomic experts following detailed guidelines provided by the Bern Secretariat. The four assessment parameters for species include its *distribution range* (Ran), *population size* (Pop), the quality and quantity of the *available habitat* for the species (H4s), and its *future prospects* (Fpr) based on the balance of pressures and conservation measures. All conclusions are evaluated in a simple "traffic light system" using one of the following four *conservation status* categories: 

*  **FV**: Favourable (green), 
*  **U1**: Unfavourable - inadequate (amber), 
*  **U2**: Unfavourable - bad (red),
*  **XX**: Unknown (grey).

### Species assessments

@tbl-conclusions-sp-fgrp provides an overview of the overall conclusions across the major species groups, whereas @tbl-conclusions-sp-para shows the distribution of the partial conclusions for the individual parameters across all the species.   

```{r}
#| label: tbl-conclusions-sp-fgrp
#| tbl-cap: "Distribution of overall conclusions across the different groups of features (FV: Favourable; U1: Unfavourable - inadequate; U2: Unfavourable - bad; XX: Unknown)"
#| warning: false
#| echo: false

lov <- stcodes %>% filter(id=="rd_ConclusionsAssessment") %>% select(sc, sl) %>% deframe

### conclusions per species groups
tmp1a <- dat$sp %>%
  filter(t_ass=="full") %>% select(Codes= E1a.0) %>% 
  group_by(Codes) %>% summarise(n=n()) 
tmp1b <- dat$sp %>% 
  filter(t_ass=="full") %>% select(fgrp, Codes= E1a.0) %>% 
  group_by(fgrp, Codes) %>% summarise(n=n()) %>% ungroup
tmp1 <- tmp1a %>%
  mutate(fgrp="all") %>% bind_rows(tmp1b, .) %>% 
  pivot_wider(names_from= Codes, values_from= n) %>%
  mutate(main= str_c(fgrps_en[fgrp], " (",fgrp,")")) %>%
  h_ensure_cols(names(lov)) %>% select(!!!c("main", names(lov)))

tmp1 %>% gt() %>% 
  cols_label(main= "Species groups") %>%
  cols_align("center") %>% cols_align("left", columns= main) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```

```{r}
#| label: tbl-conclusions-sp-para
#| tbl-cap: "Distribution of partial conclusions over the different parameters (FV: Favourable; U1: Unfavourable - inadequate; U2: Unfavourable - bad; XX: Unknown)"
#| warning: false
#| echo: false

para_sp  <- c(Ran="B7a", Pop="A8a", H4s="C5a", Fpr="D4d.0")

tmp2a <- para_sp %>%
  imap(\(ff, i) dat$sp %>% filter(t_ass=="full") %>% mutate(para=i, Codes= !!rlang::sym(ff)) %>%
    group_by(para, Codes) %>% summarise(n=n(), .groups="drop")) %>%
  list_rbind %>%
  pivot_wider(names_from= Codes, values_from= n) %>%
  mutate(main= para0_en[para]) %>%
  h_ensure_cols(names(lov)) %>% select(!!!c("main", names(lov)))
tmp2b <- para_sp %>%   # needed only by the figures below
  imap(\(ff, i) dat$sp %>% filter(t_ass=="full") %>% mutate(fgrp, para=i, Codes= !!rlang::sym(ff)) %>%
    group_by(fgrp, para, Codes) %>% summarise(n=n(), .groups="drop")) %>%
  list_rbind

tmp2a  %>% gt() %>% 
  cols_label(main= "Assessment parameters") %>%
  cols_align("center") %>% cols_align("left", columns= main) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```


Below we also visualise the distribution of the conclusions across the different species groups and assessment parameters (@fig-sp-conclusions). 

:::{#fig-sp-conclusions layout="[[1, 1],[1,1,1,1]]"}
```{r}
#| label: conclusions_1a
#| warning: false
#| echo: false

# the overall-overall conclusions (pie) 
tmp1a %>% #filter(!is.na(Codes)) %>% 
  {ggplot(.) + geom_bar(aes(x="", y=n, fill=Codes), stat="identity", width=1, color="white") +
  scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
  coord_polar(theta="y", direction=-1) + 
  theme_void(base_size= 16)}

```

```{r}
#| label: conclusions_1b
#| warning: false
#| echo: false

# the overall conclusions per species group (bars) 
tmp1b %>% #filter(!is.na(Codes)) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + #, width=1, color="white") +
  scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
  theme_minimal(base_size= 16) + theme(axis.title = element_blank(), legend.position= "none")
  }

```


```{r}
#| label: conclusions_2a
#| warning: false
#| echo: false

para1= "Ran" 

tmp2a %>% 
  mutate(fgrp=fct("all", names(fgrps_en))) %>% bind_rows(tmp2b, .) %>%
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
    scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: conclusions_2b
#| warning: false
#| echo: false

para1= "Pop" 

tmp2a %>% 
  mutate(fgrp=fct("all", names(fgrps_en))) %>% bind_rows(tmp2b, .) %>%
  filter(para==para1) %>%
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
    scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: conclusions_2c
#| warning: false
#| echo: false

para1= "H4s" 

tmp2a %>% 
  mutate(fgrp=fct("all", names(fgrps_en))) %>% bind_rows(tmp2b, .) %>%
  filter(para==para1) %>%
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
    scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: conclusions_2d
#| warning: false
#| echo: false

para1= "Fpr" 

tmp2a %>% 
  mutate(fgrp=fct("all", names(fgrps_en))) %>% bind_rows(tmp2b, .) %>%
  filter(para==para1) %>%
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
    scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

The conclusions of the species assessments in the 2025 reporting of Norway to the Bern Convention (Resultion 8). Top row: overall conclusions for all species (pie chart) and for major species groups (columns). Bottom row: partial conclusions for the four main assessment parameters. 
:::

### Habitat type assessments

@tbl-conclusions-ha-para and @fig-ha-conclusions provide an overview of the different conclusions for the habitat types assessed in 2025 in Norway.     

```{r}
#| label: tbl-conclusions-ha-para
#| tbl-cap: "Distribution of partial conclusions over the different parameters (FV: Favourable; U1: Unfavourable - inadequate; U2: Unfavourable - bad; XX: Unknown)"
#| warning: false
#| echo: false

# stcodes %>% filter(id=="rd_ConclusionsAssessment") %>% {.$sl}

para_ha  <- c(Ran="B7a", Are="A6a", SnF="C5a.0", Fpr="D4d.0", OvC="E1a.0")

tmp2a <- para_ha %>%
  imap(\(ff, i) dat$ha %>% filter(t_ass=="full") %>% mutate(para=i, Codes= !!rlang::sym(ff)) %>%
    group_by(para, Codes) %>% summarise(n=n(), .groups="drop")) %>%
  list_rbind 
tmp2 <- tmp2a %>% 
  pivot_wider(names_from= Codes, values_from= n) %>%
  mutate(main= para0_en[para]) %>%
  h_ensure_cols(names(lov)) %>% select(!!!c("main", names(lov)))

tmp2  %>% gt() %>% 
  cols_label(main= "Assessment parameters") %>%
  cols_align("center") %>% cols_align("left", columns= main) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```


:::{#fig-ha-conclusions layout="[[1, 1]]"}
```{r}
#| label: conclusions_ha1
#| warning: false
#| echo: false

# the overall-overall conclusions (pie) 
dat$ha %>%
  filter(t_ass=="full") %>% select(Codes= E1a.0) %>% 
  group_by(Codes) %>% summarise(n=n()) %>% 
  {ggplot(.) + geom_bar(aes(x="", y=n, fill=Codes), stat="identity", width=1, color="white") +
  scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
  coord_polar(theta="y", direction=-1) + 
  theme_void(base_size= 16)}
```

```{r}
#| label: conclusions_ha2
#| warning: false
#| echo: false

# the overall conclusions per sub-assessment (bars) 
tmp2a %>% filter(para!="OvC") %>% 
  mutate(para=fct(para, names(para0_en))) %>%
  {ggplot(.) + geom_bar(aes(x=para, y=n, fill=Codes), stat="identity", position="fill") + #, width=1, color="white") +
  scale_fill_manual(values= c(FV="yellowgreen",U1="gold2",U2="darkorange",XX="ivory3")) +
  theme_minimal(base_size= 16) + theme(axis.title = element_blank(), legend.position= "none")
  }

```

The conclusions of habitat type assessments in the 2025 reporting of Norway to the Bern Convention (Resultion 8). Left: overall conclusions for all habitat types (pie chart); right: partial conclusions for the four main assessment parameters (Ran: range, Are: area, SnF: structure and function, Fpr: future prospects)
:::

## Changes and trends

The assessment protocol offers several ways to describe past, present and future trends in the conservation status of the studied species and habitats. This includes the assessment of the *short-term* and *long-term* trends by the experts participating in the 2025 assessment, as well as a direct comparisons of the consecutive assessments (e.g comparing the conservation status of the same species & habitats assessed in 2019 and 2025). In the context of the Bern Convention *short-term trends* always refer to the past 12 years (i.e. 2013-2024 in this case), whereas *long term trends* are referring to the 24 year period before the assessment date (i.e. 2001-2024). In addition, the parameter *future prospects* discussed in the previous section also addresses trends, synthesizing expectations for changes in the other three parameters during the forthcoming 12 years (i.e. 2025-2036). 

In the next few sections we provide an overview of the main current and past trends relying on the different elements of the Bern assessment protocol.    

### Short term trends

Short term trends are first assessed individually for the different state parameters (range, population size, and habitat availability for species, and range, area, and structure/function for habitat types), and then the directions of these elementary trends are aggregated into an *overall trend qualifier*, which is the second flagship output of Bern Convention assessments, providing critical insights for the interpretation of the *overall conservation status*. In @fig-trend-qualifiers we provide an overview of these trend qualifiers for the different groups of features.     

:::{#fig-trend-qualifiers layout="[[1, 1],[8,7,3,7,3]]"}
```{r}
#| label: qualifiers_1a
#| warning: false
#| echo: false

lov <- stcodes %>% filter(id=="rd_ConclusionsTrend") %>% select(sc,sl) %>% deframe

### overall trend qualifiers per species groups
tmp1 <- dat$sp %>%
  filter(t_ass=="full") %>% select(Codes= E2a.0) %>% 
  bind_rows(dat$ha %>% filter(t_ass=="full") %>% select(Codes= E2a.0)) %>%
  mutate(Codes=Codes %>% fct(names(lov))) %>%
  group_by(Codes) %>% summarise(n=n()) 

# the overall trend qualifiers (pie) 
tmp1 %>% 
  {ggplot(.) + geom_bar(aes(x="", y=n, fill=Codes), stat="identity", width=1, color="white") +
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
  coord_polar(theta="y", direction=-1) + 
  theme_void(base_size= 16)}

```

```{r}
#| label: qualifiers_1b
#| warning: false
#| echo: false

tmp2 <- dat$sp %>%
  filter(t_ass=="full") %>% select(fgrp, Codes= E2a.0) %>% 
  bind_rows(dat$ha %>% filter(t_ass=="full") %>% select(fgrp, Codes= E2a.0)) %>%
  mutate(Codes=Codes %>% fct(names(lov))) %>%
  group_by(fgrp, Codes) %>% summarise(n=n()) 

# the overall conclusions per species group (bars) 
tmp2 %>% #filter(!is.na(Codes)) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + #, width=1, color="white") +
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
  theme_minimal(base_size= 16) + theme(axis.title = element_blank(), legend.position= "none")
  }

```


```{r}
#| label: qualifiers_2_ran_calc
#| warning: false
#| echo: false
#| fig-width: 8     
#| fig-height: 4.5  

para_sp  <- c(Ran="B3a", Pop="A4a", H4s="C3a")
para_ha  <- c(Ran="B3a", Are="A2a", SnF="C3a")
lov <- stcodes %>% filter(id=="rd_ConclusionsTrend") %>% select(sc,sl) %>% deframe

tmp2a <- para_sp %>%
  imap(\(ff, i) dat$sp %>% filter(t_ass=="full") %>% mutate(para=i, Codes= !!rlang::sym(ff)) %>%
    group_by(fgrp, para, Codes) %>% summarise(n=n(), .groups="drop")) %>%
  list_rbind
tmp2b <- para_ha %>%
  imap(\(ff, i) dat$ha %>% filter(t_ass=="full") %>% mutate(para=i, Codes= !!rlang::sym(ff)) %>%
    group_by(fgrp, para, Codes) %>% summarise(n=n(), .groups="drop")) %>%
  list_rbind
tmp2 <- bind_rows(tmp2a, tmp2b) %>%
  mutate(Codes= Codes %>% str_sub(1,1) %>%  fct(names(lov))) 
# tmp2$Codes %>% table

para1= "Ran" 

tmp2 %>% 
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: qualifiers_2_pop
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 4.5  

para1= "Pop" 

tmp2 %>% 
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: qualifiers_2_are
#| warning: false
#| echo: false
#| fig-width: 3
#| fig-height: 4.5  

para1= "Are" 

tmp2 %>% 
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: qualifiers_2_h4s
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 4.5  

para1= "H4s" 

tmp2 %>% 
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
```

```{r}
#| label: qualifiers_2_snf
#| warning: false
#| echo: false
#| fig-width: 3
#| fig-height: 4.5  

para1= "SnF" 

tmp2 %>% 
  filter(para==para1) %>% 
  {ggplot(.) + geom_bar(aes(x=fgrp, y=n, fill=Codes), stat="identity", position="fill") + 
  scale_fill_manual(name="Trend direction", labels=lov, 
                    values= c(I="yellowgreen",S="gold2",D="darkorange",U="ivory3")) +
    labs(title=para1) + theme_minimal(base_size= 32) + 
    theme(axis.title= element_blank(), legend.position= "none")
    } %>% plot
# dev.size("in")
```

Overall short-term (2013-2024) trends identified during the 2025 assessments. Top row: the overall "*trend qualifiers*" among all assessments (pie chart) and the distribution of trend types across the main groups of features (columns). Bottom row: short term trends for the main assessment parameters. (ma: mammals, fa= fishes & amphibians, im= insects and molluscs, vp: vascular plants, mo: mosses, ht: habitat types; Ran: range, Pop: population, Are: area, H4s: habitat for species, SnF: structure and function)
:::

### Long term trends 

In this reporting round, *long-term trends* were only assessed for some of the parameters (range & population size for species, and range and area for habitat types), and only in cases when the expert were long-term evidence was available. These cases  means that...  ... @tbl-xxxx provides an overview of the 


### Changes since 2019

The 2019 reporting period was the first reporting period for Norway under Resolution 8 of the Bern Convention. This reporting period was considered to be a pilot exercise by the Bern Convention Secretariat, so only very few species and habitat types were selected for reporting. @tbl-changes-19 provides an overview of the changes in the overall assessment conclusions for the overlapping species.

```{r}
#| label: tbl-changes-19
#| tbl-cap: "Changes in the overall conclusions of the repeated assessments between 2019 and 2025 (FV: Favourable; U1: Unfavourable - inadequate; U2: Unfavourable - bad; XX: Unknown)"
#| warning: false
#| echo: false

### conclusions in 2019
tmp1a <- i2019all$sr_re %>%
  select(b_id=fea, bgr, cc19= sr_conclusion_assessment_v, co19= sr_conclusion_assessment_trend_v) %>%
  filter(cc19!="")  # here I also pull out the old trend qualifiers (co19) but then I don't use them... 
tmp1 <- i2019all$hr_re %>%
  select(b_id=fea, bgr, cc19= hr_conclusion_assessment_v, co19= hr_conclusion_assessment_trend_v) %>%
  filter(cc19!="") %>%
  bind_rows(tmp1a)
# tmp1$cc19 %>% table #%>% dput

tmp2a <- dat$sp %>%
  filter(t_ass=="full") %>% 
  select(b_id, bgr, rl_navn, fgrp, cc25= E1a.0, co25= E2a.0) 
tmp2 <- dat$ha %>%
  filter(t_ass=="full") %>% 
  select(b_id=b_id, bgr, rl_navn, fgrp, cc25= E1a.0, co25= E2a.0) %>%
  bind_rows(tmp2a) %>%
  inner_join(tmp1, by= join_by(b_id, bgr)) %>%
  mutate(cc= str_c(cc19, " --> ", cc25)) %>%
  group_by(cc, rl_navn) %>% summarise(bgrs= str_c(bgr, collapse=", "), N=n()) %>%
  summarise(N=sum(N), assmts= str_c(rl_navn," (",bgrs,")",collapse="; ") %>% h_cap1st) %>%
  arrange(desc(N))

tmp2 %>% gt() %>% 
  cols_label(cc= "Change", assmts= "Assessments") %>%
  cols_align("center") %>% cols_align("left", columns= assmts) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))
```

If the overall conservation status of a species/habitat changed since 2019 then the experts were were requested to assess the main reasons for the change. They could select more then one reason from a list that included *genuine change*, changes in *data* or *methods*. The results are summarised in @tbl-change-reasons-19.

```{r}
#| label: tbl-change-reasons-19
#| tbl-cap: "The main reasons for changes in the overall conclusions since 2019"
#| warning: false
#| echo: false

# fields <- c(genuine="E4a01", Pop="A4a", H4s="C3a")
lov <- stcodes %>% filter(id=="rd_ChangeReason_Main") %>% select(sc,sl) %>% mutate(sl= h_cap1st(sl)) %>% deframe

tmp2a <- dat$sp %>%
  filter(t_ass=="full") %>% 
  select(b_id, bgr, rl_navn, fgrp, starts_with("E4a")) 
tmp2 <- dat$ha %>%
  filter(t_ass=="full") %>% 
  select(b_id, bgr, rl_navn, fgrp, starts_with("E4a")) %>%
  bind_rows(tmp2a) %>%
  pivot_longer(starts_with("E4a"), values_to="cc", values_drop_na=T) %>%
  group_by(cc, rl_navn) %>% summarise(bgrs= str_c(bgr, collapse=", "), N=n()) %>%
  summarise(N=sum(N), assmts= str_c(rl_navn," (",bgrs,")",collapse="; ") %>% h_cap1st) %>%
  mutate(cc= lov[cc]) # %>%
  # arrange(desc(N))

tmp2 %>% gt() %>% 
  cols_label(cc= "Reasons for status change between 2019 & 2025", assmts= "Assessments") %>%
  # cols_align("center") %>% cols_align("left", columns= c(cc, assmts)) %>%
  sub_missing() %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))
```

## Background factors 

A major element of the Bern Convention (Resolution 8) reporting protocol is the assessment of *pressures* (i.e. threats) and *conservation measures* affecting the status and trends of the selected species and habitat types. These factors have a key role in the assessment of the *future prospects* parameter, but and they can also provide relevant information by  themselves, e.g. for effective conservation planning. The assessment protocol offers a long list of predefined types of pressures and conservation measures, and the experts could choose from these lists, assigning an ordered list of *pressures* and *measures* to each species and habitat type. 
<!-- In addition, the experts also characterised each pressure and measure from the perspective of several further properties, specifying e.g. if a pressure is an ongoing or only an expected future threat, and approximately what portion of the population is affected by it.  -->

### Key pressures

Beyond just listing the relevant pressures the experts also had to characterise them from three different perspectives: its timing (past, ongoing or expected future pressure), extent (the share of the population/area it affects), and influence (the degree of impact on the affected population/area). @tbl-press-list presents an overview of the pressures that were most commonly listed as "ongoing" threats across all species and habitat types. 

```{r}
#| label: tbl-press-list
#| tbl-cap: "The ten most most commmon active (ongoing) pressures identified during the assessments (N: number of assessments where the given pressure was identified; ma: mammals, fa= fishes & amphibians, im= insects and molluscs, vp: vascular plants, mo: mosses, ht: habitat types)"
#| warning: false
#| echo: false

tmp1 <- dat$sp %>% 
  select(ftid, fgrp, rl_navn, bgr, starts_with("D1a"), D1b01, D1c) %>%  #D1b01, D1c: to bring back the IAS 
  bind_rows(select(dat$ha, ftid, fgrp, bgr, starts_with("D1a"), D1b01, D1c)) %>%
  mutate(D1a21= if_else(is.na(D1b01),NA,"PI01"), D1a21.2=if_else(is.na(D1b01),"","pres"),
         D1a22= if_else(is.na(D1c),NA,"PI02"), D1a22.2= if_else(is.na(D1c),"","pres")) %>% 
  select(-D1b01,-D1c) %>% #TODO: next time keep these codes in the list :(
  pivot_longer(starts_with("D1a")) %>%
  separate_wider_delim(name, ".", names=c("n_tmp", "subcol"), too_few="align_start") %>%
  mutate(subcol= subcol %>% replace_na("1") %>% str_c("s",.)) %>%
  pivot_wider(names_from=subcol, values_from=value) %>%
  filter(!is.na(s1)) 

# tmp1 %>% 
#   filter(str_detect(s2,"pres")) %>% 
#   group_by(s1) %>% summarise(N=n()) %>% 
#   arrange(desc(N)) %>% 
#   mutate(`Pressure name`= h_lkp(s1, pmcodes)) %>%
#   select(Code=s1, `Pressure name`, N) %>%
#   slice(1:20) %>%

# Create a custom split html label for the nanoplot column using flexbox
nano_label_features <- html("
  <div style='text-align: center; border-bottom: 1px solid #ccc; font-size: 12px;  
  padding-bottom: 3px; margin-bottom: 3px;'>
    Feature groups 
  </div>
  <div style='width: 90px; margin: 0 auto; display: flex; font-size: 8px; font-weight: normal; line-height: 1;'>
    <div style='flex: 1; text-align: center; min-width: 0;'>ma</div>
    <div style='flex: 1; text-align: center; min-width: 0;'>fa</div>
    <div style='flex: 1; text-align: center; min-width: 0;'>im</div>
    <div style='flex: 1; text-align: center; min-width: 0;'>vp</div>
    <div style='flex: 1; text-align: center; min-width: 0;'>mo</div>
    <div style='flex: 1; text-align: center; min-width: 0;'>ht</div>
  </div>")

tmp0 <- fgrps_en %>% {.[-length(.)]} %>% names %>% set_names
tmp1 %>%
  filter(str_detect(s2,"pres")) %>% 
  group_by(s1) %>% summarise(N=n(), fgrps= list(map_int(tmp0, \(x) sum(fgrp==x)))) %>% 
  arrange(desc(N)) %>%
  mutate(label= h_lkp(s1, pmcodes)) %>%
  relocate(s1, label) %>% slice(1:20) %>%
  gt() %>%
  sub_missing() %>%
  cols_nanoplot(columns= fgrps, plot_type= "bar", 
    options= nanoplot_options(data_bar_fill_color= "steelblue", data_bar_stroke_color= "transparent")) %>%
  cols_label(s1="Code", label="Pressure name", nanoplots=nano_label_features) %>%
  cols_align("center") %>% cols_align("left", columns= label) %>%
  cols_width(label ~ pct(80)) %>% # to make this column greedy
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```


### Main conservation measures

Similarly to the pressures, the experts also identified the morst important conservation measures for each feature in each biogeographic region. This included all "ongoing" (actively implemented) conservation measures, as well as any clearly useful, but not yet implemented measures. In addition, for each already implemented measure the experts also documented if the measures were implemented only within the Emerald sites, or also beyond them. @tbl-meas-list presents an overview of the most frequently applied conservation measures. 

```{r}
#| label: tbl-meas-list
#| tbl-cap: "The ten most frequently implemented conservation measures, as identified during the assessments (N: number of assessments where the given measure was identified; ma: mammals, fa= fishes & amphibians, im= insects and molluscs, vp: vascular plants, mo: mosses, ht: habitat types)"
#| warning: false
#| echo: false

tmp1.meas <- dat$sp %>% 
  select(ftid, fgrp, rl_navn, bgr, starts_with("D2c")) %>%   
  bind_rows(select(dat$ha, ftid, fgrp, bgr, starts_with("D2c"))) %>%
  pivot_longer(starts_with("D2c")) %>%
  separate_wider_delim(name, ".", names=c("n_tmp", "subcol"), too_few="align_start") %>%
  mutate(subcol= subcol %>% replace_na("1") %>% str_c("s",.)) %>%
  pivot_wider(names_from=subcol, values_from=value) %>%
  filter(!is.na(s1)) 

# nano_label_features: this is recycled from the previous chunk
tmp0 <- fgrps_en %>% {.[-length(.)]} %>% names %>% set_names
tmp1.meas %>%
  filter(s4=="T") %>%
  group_by(s1) %>% summarise(N=n(), fgrps= list(map_int(tmp0, \(x) sum(fgrp==x)))) %>% 
  arrange(desc(N)) %>%
  mutate(label= h_lkp(s1, pmcodes)) %>%
  relocate(s1, label) %>% slice(1:10) %>%
  gt() %>%
  sub_missing() %>%
  cols_nanoplot(columns= fgrps, plot_type= "bar", 
    options= nanoplot_options(data_bar_fill_color= "steelblue", data_bar_stroke_color= "transparent")) %>%
  cols_label(s1="Code", label="Conservation measure name", nanoplots=nano_label_features) %>%
  cols_align("center") %>% cols_align("left", columns= label) %>%
  cols_width(label ~ pct(80)) %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```


### Role of the Emerald Network

Protected areas, and in particular the Emerald Network is assumed to provide substantial support for endangered species and habitats. Several parts of the Bern assessment explore the influence of the Emerald Network, thus providing useful feedback for conservation management. In @tbl-eme-trends we compare *within Emerald* short-term trends of several parameters with the general trends of the same parameters within the entire biogeographic region. This comparison is only sensitive to the direction of the trend, so significant differences between slight and drastic declines (or improvements) can still be marked as "same within/beyond Emerald" in @tbl-eme-trends.      


```{r}
#| label: tbl-eme-trends
#| tbl-cap: "Comparison of short term trends in several parameters within the Emerald network and outside of it." 
#| warning: false
#| echo: false

para <- list(Pop=c("A4a","A4c"), H4s=c("C3a","C3b"), Are= c("A2a","A2c"), SnF=c("C3a","C3b"))
tmp <-  c(Pop="sp", H4s="sp", Are="ha", SnF="ha")
lov <-  c(better= "Better *within* Emerald", same= "Same trend within/beyond", 
          worse=  "Better *beyond* Emerald", unk=  "Unknown (at least partly)") 

tmp2 <- para %>%
  imap(\(x,i) dat[[tmp[i]]] %>% filter(t_ass=="full") %>% 
         mutate(c_reg= !!rlang::sym(x[1]), c_eme= !!rlang::sym(x[2]))) %>%
  map( \(x)   x %>% mutate(c_eme=str_sub(c_eme,1,1), c_reg=str_sub(c_reg,1,1))) %>%
  imap(\(x,i) x %>% mutate(para=i, cc= str_c(c_eme, c_reg))) %>%
  map( \(x)   x %>% group_by(para, cc) %>% summarise(N=n(), spp= list(sort(unique(rl_navn))), .groups="drop")) %>%
  list_rbind %>%
  # mutate(cc= if_else(str_detect(cc, "U"), NA, cc)) %>%
  mutate(cc2= case_match(cc, c("SD","ID","IS") ~ "better",
                             c("DD","SS","II") ~ "same",
                             c("DS","DI","SI") ~ "worse")) %>%
  mutate(cc2= replace_na(cc2, "unk") %>% fct(names(lov))) %>%
  group_by(para, cc2) %>% summarise(N=sum(N), spp=list(sort(unique(list_c(spp)))), .groups="drop") %>%
  arrange(cc2)
tmp2b <- tmp2 %>%
  select(-spp) %>%
  pivot_wider(names_from= para, values_from= N) %>% 
  relocate(!!!c("cc2", names(para0_en[names(tmp)]))) %>%
  mutate(cc2= lov[cc2])

tmp2b %>% gt() %>% 
  cols_label(.list= c(cc2="Trends within/beyond Emerald", para0_en[names(tmp)])) %>%
  cols_align("center") %>% cols_align("left", columns= cc2) %>%
  sub_missing() %>%
  fmt_markdown(columns= cc2) %>% #!!!
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```

```{r}
#| label: txt-eme-trends-diag
#| warning: false
#| echo: true

# A quick temporary overview of the most interesting cases of the table above:
tmp2 %>% 
  filter(para %in% c("Pop", "H4s"), cc2 %in% c("better", "worse")) %>%
  mutate(species=map_chr(spp, \(x) str_c(x, collapse=", "))) %>%
  select(para, cc2, species) %>%
  print

```


In addition we can also compare the implementation of the conservation measures within and beyond the Emerald Network based on the answers of the experts. @tbl-eme-meas offers such a comparison. 

```{r}
#| label: tbl-eme-meas
#| tbl-cap: "The location of the most frequently implemented conservation measures: predominantly within Emerald sites (N~in~), mostly outside of the Emerald Network (N~out~), or both within and beyond Emerald (N~i.o~)"
#| warning: false
#| echo: false

tmp0 <- c(i="inside", io="both", o="outside")
tmp1.meas %>%
  filter(s4=="T") %>%  # "T" means taken
  group_by(s1) %>% summarise(N=n(), tmp=list(map(tmp0, \(x) sum(s2==x)))) %>% 
  mutate(tmp= map(tmp, as_tibble)) %>% unnest(tmp) %>% 
  arrange(desc(N)) %>%
  mutate(label= h_lkp(s1, pmcodes)) %>%
  relocate(s1, label) %>% slice(1:10) %>% 
  select(-N) %>%
  gt() %>%
  sub_missing() %>%
  cols_label(s1="Code", label="Conservation measure name", #N=md("N~tot~"), 
             i=md("N~in~"), io=md("N~i.o~"), o=md("N~out~")) %>%
  cols_align("center") %>% cols_align("left", columns= label) %>%
  # cols_width(label ~ pct(80)) %>%
  tab_style(style = cell_text(weight= "bold"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(v_align= "top"), locations = cells_body()) %>%
  tab_options(table.width = pct(100), data_row.padding = px(3))

```

## Fjellrev: a success story

TBC...


## Data availability

(?) TBD/TBC...


## References


## Annexes

* A table with all species & habitats (incl latin names) and their main conclusions
* (??) A list of simple “conclusion figures” for each species & habitat (one per page)
